---
title: "QA instructions"
date: "February 23, 2018"
output:
  md_document:
    variant: markdown_github
---

## Set up

```{r setup, eval=FALSE}
install.packages("remotes")
remotes::install_github("rstudio/websocket")
```

## Creating a websocket
First, create a websocket client by running:

```{r createWS}
# Create a websocket using the websocket.org test server
ws <- websocket::WebsocketClient$new("ws://echo.websocket.org/",
  onMessage = function(msg) {
    cat("Client got msg: ", msg, "\n")
  },
  onDisconnected = function() {
    cat("Client disconnected\n")
  }
)
```

For now, this displays an error with the red message:

```
easywsclient: connecting: host=echo.websocket.org port=80 path=/
Connected to: ws://echo.websocket.org/
```

This is an unfortunate side-effect from the underlying c++ library we're using, but it doesn't affect the functionality.

## Sending and receiving messages

Then, you can send an arbitrary number of messages like so:

```{r}
ws$send("hello")
```

For each message `msg` you send, the following should be printed to the R console: "Client got msg: `msg`".

## Closing the websocket

Finally, when you're done, close the websocket:

```{r}
ws$close()
```

When you do this, the following should be printed to the R console: "Client disconnected"